---
title: 'Project #3: Escapades in Market Risk'
subtitle: "Metal market ETF analysis"
output:
  html_document:
    toc: true
    toc_float: true
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[CO,CE]{\thetitle}
- \fancyfoot[CO,CE]{Copyright 2018, William G. Foote}
- \fancyhead[RE,RO]{\thepage}
- \renewcommand{\footrulewidth}{0.5pt}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, warning=FALSE, message=FALSE)
knitr::opts_chunk$set(tidy = TRUE)
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=36))
knitr::opts_chunk$set(size = "small")
knitr::opts_hooks$set(fig.width = function(options) {
  if (options$fig.width < options$fig.height) {
    options$fig.width = options$fig.height
  }
  options
})
knitr::knit_hooks$set(mysize = function(before, options, envir) {
  if (before) 
    return(options$size)
})
```

## Purpose, Process, Product

Various `R` features and finance topics will be reprised in this chapter. Specifically we will practice reading in data, exploring time series, estimating auto and cross correlations, and investigating volatility clustering in financial time series. We will build a simple financial web application. We will explore the extremes of distributions of financial returns. We will summarize our experiences in debrief. 

## Assignment

This assignment will span Live Sessions 3 and 4 (two weeks). The project (3) is due before Live Session 5. Submit into **Coursework > Assignments and Grading > Project 4 > Submission** an `RMD`  file with filename **lastname-firstname_Project4.Rmd**. If you have difficulties submitting a `.Rmd` file, then submit a `.txt` file. Also submit an `html` file with filename **lastname-firstname_Project3.html**. 

1. Use headers (##), r-chunks for code, and text to build a report that addresses the two parts of this project.

2. List in the text the 'R' skills needed to complete this project.

3. Explain each of the functions (e.g., `ggplot()`) used to compute and visualize results.

4. Discuss how well did the results begin to answer the business questions posed at the beginning of each part of the project.

## Flexdashboard

We begin to expand our range of production capabilities. we started with an `R` script in project 1, then moved onto a `R Markdown` PDF file in projects 2 and 3. Now we introduce ourselves to the `flexd8ashboard` package with a little bit of `shiny` to get us used to building a web application directly.

1. Install the `flexdashboard` and `shiny` packages. In the Rstudio console type `library(c("flexdashboard", "shiny"))` on the command line to attach these packages to the workspace.

2. Go to the RStudio `flexdashboard` site to learn about the basics of building a web application.

3. Go to `File` and click on `New File` where you will choose `R Markdown`. Because you already loaded `flexdashboard` this next step will work. In the dialog box choose `From Template`. You should see a `Flex Dashboard` template choice in the list box, which you will select and then click `OK`. A new `Rmd` file will then appear in the scripts pane of RStudio.

4. `Knit` this new file. A dialog box will appear for you to name the file. 

5. Start to modify this template to document your own data analysis journey from here on out. Each live session will contain a short segment on doing exactly that. Use the `extreme-2` application to guide your work.

## Problem

A freight forwarder with a fleet of bulk carriers wants to optimize their portfolio in  metals markets with entry into the nickel business and use of the tramp trade.  Tramp ships are the company's "swing" option without any fixed charter or other constraint. They allow the company flexibility in managing several aspects of freight uncertainty. The call for tramp transportation is a _derived demand_ based on the value of the cargoes. This value varies widely in the spot markets. The company allocates \$250 million to manage receivables. The company wants us to:

1.	Retrieve and begin to analyze data about potential commodities for diversification,
2.	Compare potential commodities with existing commodities in conventional metal spot markets,
3.	Begin to generate economic scenarios based on events that may, or may not, materialize in the commodities.
4.	The company wants to mitigate their risk by diversifying their cargo loads. This risk measures the amount of capital the company needs to maintain its portfolio of services.

Here is some additional detail.

1.	Product: Metals commodities and freight charters
2.	Metal, Company, and Geography:
    a. Nickel: MMC Norilisk, Russia
    b. Copper: Codelco, Chile and MMC Norilisk, Russia
    c. Aluminium: Vale, Brasil and Rio Tinto Alcan, Australia
3.	Customers: Ship Owners, manufacturers, traders
4.  All metals are traded on the London Metal Exchange. Metal ETFs are also traded on various stock exchanges.

### Key business questions

1.	How would the performance of these commodities affect the size and timing of shipping arrangements?
2.	How would the value of new shipping arrangements affect the value of our business with our current customers?
3.	How would we manage the allocation of existing resources given we have just landed in this new market?

### Getting to a reponse: more detailed questions

1. What is the decision the freight-forwarder must make? List key business questions and data needed to help answer these questions and support the freight-forwarder's decision. Retrieve data and build financial market detail into the data story behind the questions.

2. Develop the stylized facts of the markets the freight-forwarder faces. Include times series plots. Calculate and display in a table the summary statistics, including quantiles, of each of these series. Use autocorrelation, partial autocorrelation, and cross correlation functions to understand some of the persistence of returns including leverage and volatility clustering effects. Use quantile regressions to develop the distribution of sensitivity of each market to spill-over effects from other markets. Interpret these stylized "facts" in terms of the business decision the freight-forwarder makes.

3. How much capital would the freight-forwarder need? Determine various measures of risk in the tail of each metal's distribution. Then figure out a loss function to develop the portfolio of risk, and the determination of risk capital the freight-forwarder might need. Confidence intervals arond capital metrics can support a risk management plan with varying tail experience thresholds.

### Data

Use daily price data for [three metal exchange traded funds](https://etfdb.com/etfdb-category/metals/):

- JJC: copper fund

- JJN: nickel fund

- JJU: aluminium fund

You may use this or similar code to get you started on the analysis.

```{r example}
options(digits = 4, scipen = 999999)
library(psych)
library(ggplot2)
library(GGally)
library(lubridate)
library(dplyr)
library(quantreg)
library(forecast)
library(tidyquant)
library(timetk)
library(quantmod)
library(matrixStats)
library(QRM)
#
metals_env <- new.env()
symbols <- c("JJC", "JJN", "JJU")
getSymbols(symbols) #, env = stocks_env) # using quantmod
data <- JJC # COPPER
data <- data[ , 6] # only adjusted close  
colnames(data) <- "copper"
r_JJC <- diff(log(data))[-1] 
# convert xts object to a tibble or data frame
p_JJC <- data %>% as_tibble() %>% mutate(date = index(data), month = month.abb[month(index(data))])
# repeat
data <- JJN #NICKEL
data <- data[ , 6]  
colnames(data) <- "nickel"
r_JJN <- diff(log(data))[-1]
p_JJN <- data %>% as_tibble() %>% mutate(date = index(data), month = month.abb[month(index(data))])
# and again
data <- JJU
data <- data[ , 6]  
colnames(data) <- "aluminium"
r_JJU <- diff(log(data))[-1]
p_JJU <- data %>% as_tibble() %>% mutate(date = index(data), month = month.abb[month(index(data))])#rate_IYM <- data %>% mutate(diff(log(p_IYM))[-1])
# merge by date (as row name)
price <- merge(p_JJC, p_JJN)
price <- merge(price, p_JJU)
return <- merge(copper = r_JJC, nickel = r_JJN, aluminium = r_JJU, all = FALSE)
# calculute within month correlations and choose lower triangle of correlation matrix
r_corr <- apply.monthly(return, FUN = cor)[, c(2, 3, 6)]
colnames(r_corr) <- c("copper_nickel", "copper_aluminium", "nickel_aluminium")
# calculate within month standard deviations using MatrixStats
r_vols <- apply.monthly(return, FUN = colSds)
# long format ("TIDY") price tibble for possible other work
price_tbl <- price %>% as_tibble() %>% gather(k = symbol, value = price, copper, nickel, aluminium ) %>% select(symbol, date, price)
return_tbl <- price_tbl %>% group_by(symbol) %>% tq_transmute(mutate_fun = periodReturn, period = "daily", type = "log", col_rename = "daily_return") %>% mutate(abs_return = abs(daily_return))
str(return_tbl)
# 
corr_tbl <- r_corr %>% as_tibble() %>% mutate(date = index(r_corr)) %>% gather(key = assets, value = corr, -date)
vols_tbl <- r_vols %>% as_tibble() %>% mutate(date = index(r_vols)) %>% gather(key = assets, value = vols, -date) 
#
corr_vols <- merge(r_corr, r_vols)
corr_vols_tbl <- corr_vols %>% as_tibble() %>% mutate(date = index(corr_vols))
#
n <-  10000 # lots of trials, each a "day" or an "hour"
z <- rt(n, df = 30)
garch_sim_t <- function(n = 1000, df = 30, omega = 0.1, alpha = 0.8, phi = 0.05, mu = 0.01){
  n <- n # lots of trials, each a "day" or an "hour"
  # set.seed(seed)
  z <- rt(n, df = df) 
  e <-  z # store variates
  y <-  z # returns: store again in a different place
  sig2 <-  z^2 # create volatility series
  omega <-  omega #base variance
  alpha <-  alpha #vols Markov dependence on previous variance
  phi <-  phi # returns Markov dependence on previous period
  mu <-  mu # average return
  for (t in 2:n) { # Because of lag start at second
    e[t] <- sqrt(sig2[t])*z[t]           # 1. e is conditional on sig
    y[t] <-  mu + phi*(y[t-1]-mu) + e[t] # 2. generate returns
    sig2[t+1] <-  omega + alpha * e[t]^2 # 3. generate new sigma^2
    }
  return <- list(
    sim_df_vbl <- data_frame(t = 1:n, z = z, y = y, e = e, sig = sqrt(sig2)[-(n+1)] ),
    sim_df_title <- data_frame(t = 1:n, "1. Unconditional innovations" = z, "4. Conditional returns" = y, "3. Conditional innovations" = e, "2. Conditional volatility" = sqrt(sig2)[-(n+1)] )
  )
}
#
price_etf <- price %>% select(copper, nickel, aluminium) # 3 risk factors (rf)
price_0 <- as.numeric(tail(price_etf, 1))
shares <- c(60000, 75000, 50000)
price_last <- price[length(price$copper), 3:5] #(TAN, ICLN, PBW) %>% as.vector()
w <- as.numeric(shares * price_last)
return_hist <- na.omit(apply(log(price[, 3:5]), 2, diff))
# Fan these across the length and breadth of the risk factor series
weights_rf <- matrix(w, nrow=nrow(return_hist), ncol=ncol(return_hist), byrow=TRUE)
## We need to compute exp(x) - 1 for very small x: expm1 accomplishes this
loss_rf <- -rowSums(expm1(return_hist) * weights_rf)
loss_df <- data_frame(loss = loss_rf, distribution = rep("historical", each = length(loss_rf)))
#
ES_calc <- function(data, prob){
  threshold <- quantile(data, prob)
  result <- mean(data[data > threshold])
}

#
n_sim <- 1000
n_sample <- 100
prob <- 0.95
ES_sim <- replicate(n_sim, ES_calc(sample(loss_rf, n_sample, replace = TRUE), prob))
#
summary(ES_sim)
#
# mean excess plot to determine thresholds for extreme event management
data <- as.vector(loss_rf) # data is purely numeric
umin <-  min(data)         # threshold u min
umax <-  max(data) - 0.1   # threshold u max
nint <- 100                # grid length to generate mean excess plot
grid_0 <- numeric(nint)    # grid store
e <- grid_0                # store mean exceedances e
upper <- grid_0            # store upper confidence interval
lower <- grid_0            # store lower confidence interval
u <- seq(umin, umax, length = nint) # threshold u grid
alpha <- 0.95                  # confidence level
for (i in 1:nint) {
    data <- data[data > u[i]]  # subset data above thresholds
    e[i] <- mean(data - u[i])  # calculate mean excess of threshold
    sdev <- sqrt(var(data))    # standard deviation
    n <- length(data)          # sample size of subsetted data above thresholds
    upper[i] <- e[i] + (qnorm((1 + alpha)/2) * sdev)/sqrt(n) # upper confidence interval
    lower[i] <- e[i] - (qnorm((1 + alpha)/2) * sdev)/sqrt(n) # lower confidence interval
  }
mep_df <- data.frame(threshold = u, threshold_exceedances = e, lower = lower, upper = upper)
loss_excess <- loss_rf[loss_rf > u] - u
quantInv <- function(distr, value) ecdf(distr)(value)
u_prob <- quantInv(loss_rf, 200000)
ES_mep <- mean(loss_rf[loss_rf > u_prob])
#u_prob
#ES_mep
```

