---
title: "The answer is always 42"
output:
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
---

```{r setup, include=FALSE}
rm(list = ls())

library(ggplot2)
library(flexdashboard)
library(shiny)
library(QRM)
library(qrmdata)
library(xts)
library(zoo)
library(psych)

# PAGE: Exploratory Analysis
data <- na.omit(read.csv("data/metaldata.csv", header = TRUE))
prices <- data
# Compute log differences percent using as.matrix to force numeric type
data.r <- diff(log(as.matrix(data[, -1]))) * 100
# Create size and direction
size <- na.omit(abs(data.r)) # size is indicator of volatility
#head(size)
colnames(size) <- paste(colnames(size),".size", sep = "") # Teetor
direction <- ifelse(data.r > 0, 1, ifelse(data.r < 0, -1, 0)) # another indicator of volatility
colnames(direction) <- paste(colnames(direction),".dir", sep = "")
# Convert into a time series object: 
# 1. Split into date and rates
dates <- as.Date(data$DATE[-1], "%m/%d/%Y")
dates.chr <- as.character(data$DATE[-1])
#str(dates.chr)
values <- cbind(data.r, size, direction)
# for dplyr pivoting and ggplot2 need a data frame also known as "tidy data"
data.df <- data.frame(dates = dates, returns = data.r, size = size, direction = direction)
data.df.nd <- data.frame(dates = dates.chr, returns = data.r, size = size, direction = direction, stringsAsFactors = FALSE) 
#non-coerced dates for subsetting on non-date columns
# 2. Make an xts object with row names equal to the dates
data.xts <- na.omit(as.xts(values, dates)) #order.by=as.Date(dates, "%d/%m/%Y")))
#str(data.xts)
data.zr <- as.zooreg(data.xts)
returns <- data.xts # watch for this data below!

# PAGE: Market risk 
corr_rolling <- function(x) {	
  dim <- ncol(x)	
  corr_r <- cor(x)[lower.tri(diag(dim), diag = FALSE)]	
  return(corr_r)	
}
vol_rolling <- function(x){
  library(matrixStats)
  vol_r <- colSds(x)
  return(vol_r)
}
ALL.r <- data.xts[, 1:3]
window <- 90 #reactive({input$window})
corr_r <- rollapply(ALL.r, width = window, corr_rolling, align = "right", by.column = FALSE)
colnames(corr_r) <- c("nickel.copper", "nickel.aluminium", "copper.aluminium")
vol_r <- rollapply(ALL.r, width = window, vol_rolling, align = "right", by.column = FALSE)
colnames(vol_r) <- c("nickel.vol", "copper.vol", "aluminium.vol")
year <- format(index(corr_r), "%Y")
r_corr_vol <- merge(ALL.r, corr_r, vol_r, year)
# Market dependencies
#library(matrixStats)
R.corr <- apply.monthly(as.xts(ALL.r), FUN = cor)
R.vols <- apply.monthly(ALL.r, FUN = colSds) # from MatrixStats	
# Form correlation matrix for one month 	
R.corr.1 <- matrix(R.corr[20,], nrow = 3, ncol = 3, byrow = FALSE)	
rownames(R.corr.1) <- colnames(ALL.r[,1:3])	
colnames(R.corr.1) <- rownames(R.corr.1)	
R.corr.1
R.corr <- R.corr[, c(2, 3, 6)]
colnames(R.corr) <- c("nickel.copper", "nickel.aluminium", "copper.aluminium") 	
colnames(R.vols) <- c("nickel.vols", "copper.vols", "aluminium.vols")	
R.corr.vols <- na.omit(merge(R.corr, R.vols))
year <- format(index(R.corr.vols), "%Y")
R.corr.vols.y <- data.frame(nickel.correlation = R.corr.vols[,1], copper.volatility = R.corr.vols[,5], year = year)
nickel.vols <- as.numeric(R.corr.vols[,"nickel.vols"])	
copper.vols <- as.numeric(R.corr.vols[,"copper.vols"])	
aluminium.vols <- as.numeric(R.corr.vols[,"aluminium.vols"])

library(quantreg)
taus <- seq(.05,.95,.05)	# Roger Koenker UI Bob Hogg and Allen Craig
fit.rq.nickel.copper <- rq(log(nickel.copper) ~ log(copper.vol), tau = taus, data = r_corr_vol)	
fit.lm.nickel.copper <- lm(log(nickel.copper) ~ log(copper.vol), data = r_corr_vol)	
#' Some test statements	
ni.cu.summary <- summary(fit.rq.nickel.copper, se = "boot")

```

Introduction
=======================================================================
## Problem

A freight forwarder with a fleet of bulk carriers wants to optimize their portfolio in  metals markets with entry into the nickel business and use of the tramp trade.  Tramp ships are the company's "swing" option without any fixed charter or other constraint. They allow the company flexibility in managing several aspects of freight uncertainty. The call for tramp transportation is a _derived demand_ based on the value of the cargoes. This value varies widely in the spot markets. The company allocates \$250 million to manage receivables. The company wants us to:

1.	Retrieve and begin to analyze data about potential commodities for diversification,
2.	Compare potential commodities with existing commodities in conventional metal spot markets,
3.	Begin to generate economic scenarios based on events that may, or may not, materialize in the commodities.
4.	The company wants to mitigate their risk by diversifying their cargo loads. This risk measures the amount of capital the company needs to maintain its portfolio of services.

Here is some additional detail.

1.	Product: Metals commodities and freight charters
2.	Metal, Company, and Geography:
    a. Nickel: MMC Norilisk, Russia
    b. Copper: Codelco, Chile and MMC Norilisk, Russia
    c. Aluminium: Vale, Brasil and Rio Tinto Alcan, Australia
3.	Customers: Ship Owners, manufacturers, traders
4.  All metals are traded on the London Metal Exchange 


Explore
=======================================================================

Column {data-width=650}
-----------------------------------------------------------------------

### Chart A

```{r}
title.chg <- "Metals Market Percent Changes"
autoplot.zoo(data.xts[,1:3]) + ggtitle(title.chg) + ylim(-5, 5)
```

Column {data-width=350}
-----------------------------------------------------------------------

### Chart B

```{r}

```

### Chart C

```{r}

```

Analyze market risk
=======================================================================

Any txt you can think of....

```{r}

```

