---
title: "Topics in Financial Analytics"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data

All of the data in the topics comes from R packages or from the zipped file you can [download from here](data/data.zip).

1. Set up a working directory on your computer. Typically this is located in the user's documents directory. In this working directory you will save *.Rmd files.

2. Within the working directory, set up a data directory called `data`. This is a subdirectory of your working directory.

3. Download all of the data into the `/data` directory on your computer.  You will be accessing this data, for the most part `CSV` (comma separated values) files using, as an example, the following code.

```{r, echo = FALSE, eval = FALSE}
data_in <- read.csv("/data/metals.csv")
```

Be sure to set up your files in a directory, then set the working directory as the source file location. In that way your call to `/data/metals.csv` (for example) will connect. Otherwise you will get the following error:

```{}
Error in file(file, "rt") : cannot open the connection
In addition: Warning message:
In file(file, "rt") :
  cannot open file 'data/metals.csv': No such file or directory
```

The **fix** is to properly set your working directory. In Rstudio on the tool bar go to **Session > Set Working Directory > Source File Location**. If you have set up the `data/` folder with the data as a sub-direcctory of the directory which is your source file location, `read.csv()` will be able to connect to your data set.

# Finance tutorials

Here are several short tutorials in financial analysis. 

- [Calculating returns](HTML/e-ln-returns.html)

- [Net present value](HTML/npv-1.html)

- [Internal rate of return](HTML/irr-1.html)

- [Exchange rates](HTML/exchange-rates-1.html)

- [Expected Shortfall confidence interval estimation](HTML/sampling-ES-1.html)

- [Private placement of capital: deal structure (IRR, PV, option pricing, strategy ROI)](HTML/private-placement-1.html)

- [Equity valuation using the binomial option pricing model](HTML/equityvaluation-1.html)

# Statistics tutorials

Modern finance theory and practice benefits greatly from advances in statistics and operations research. Here is a statistics primer that we can use to refresh our understanding and use of basic concepts and models often deployed in finance.

- [Statistical thinking](https://wgfoote.github.io/stat-thinking/)

# Topic 1: R warmups for finance {.tabset}

## Week 1

### View asynchronous material for Unit 1 -- R Warmups for Finance

Supporting files are in

- [RMD](RMD/W1-RWarmUps.Rmd)

- [PDF](PDF/W1-RWarmUps.pdf)

2. Bring specific issues to the Live Session for discussion and resolution

3. Note that **cookbook-r.com** no longer has a `package` page. Revised directions are included in the week 1 RMD file above.

4. Before the Live Session, be sure to have a working directory set up with a `data` subdirectory. Be sure to have downloaded and extracted the data for the course into the the `data` subdirectory.

### Workbook session

1. Welcome: meet and greet, respond to polls, list asynch highs and lows in chat room. (5 minutes)

2. In the `R` console type `install.packages(c("shiny", "learnr", "GGally"))`. We will use these packages and the following `Rmd` file in workbook sessions. The interactive workbook is [deployed here](https://wgfoote.shinyapps.io/finalytics-livesession-1/). Copy these files into your working directory to render your local version of this session: 

- [Video introduction to workbook #1](RMD/work-1.mp4)

- [Workbook #1 learnr RMD file](RMD/live-1.Rmd)

- [Housing price data CSV file](RMD/HousePrices.csv)

3. Review and discuss asynch questions and issues. Record solutions on the wall. (15 minutes)

4. Break-out #1: Review `summary()` assignment together. Break out into separate rooms to discuss, analyze, and formulate report. Share report together. (30 minutes)

5. Break-out #2: Review `ggplot()` assignment together. Break out into separate rooms to discuss, analyze, and formulate report. Share report together. (30 minutes)

6. Debrief this session and review the next week's asynchronous material. (10 minutes)

## Week 2 

### View asynchronous material for Unit 2 -- R Data Model

In this "learn R on the job" week you will

1. Build on plotting and computational aspects of R

- Making log differences (returns and growth rates)

- Using ggplot2 to make production grade graphics with layering techniques

- Fitting distributions to data using the `gamma` distribution

2. Learn to control data flow using `if` and functions

3. Build two financial functions.

- Net Present Value and Internal Rate of Return

Supporting files are in

- [RMD](RMD/W2-RMore.Rmd)

- [PDF](PDF/W2-RMore.pdf)

### Workbook session

1. Welcome: meet and greet, respond to polls, list asynch highs and lows in chat room. (5 minutes)

2. The interactive workbook is [deployed here](https://wgfoote.shinyapps.io/finalytics-live-2/). Copy these files into your working directory to render your local version of this session:

- [Live Session #2 learnr RMD file](RMD/live-2.Rmd)

- [Power and natural gas price data CSV file](RMD/power-gas-ne.csv)

3. Review and discuss asynch questions and issues. Record solutions on the wall. (15 minutes)

4. Assignment #1: Review pivot and lookup assignment together. Break out into separate rooms to discuss, analyze, and formulate report. Share report together. (30 minutes)

5. Assignment #2: Review simulation function assignment together. Break out into separate rooms to discuss, analyze, and formulate report. Share report together. (30 minutes)

6. Debrief this session and review the Heating Oil No. 2 and next week's asynchronous material. (10 minutes)

## Project 1: Heating Oil No. 2 Analysis

Here is your first graded project for formal feedback. You can use this project to practice the many aspects of R we covered these past two weeks.

### Project 1: HO2 Analysis

HO2 = NY Harbor No. 2 Heating Oil prices
 
- Here is a [description of the contract](http://www.cmegroup.com/trading/energy/refined-products/heating-oil_contract_specifications.html) at the NYMEX (CME Group)

- More information on HO2 is deposited [here](https://en.wikipedia.org/wiki/Heating_oil)

- Fuel oil is used to hedge a variety of refined oil products including Jet-A (Kerosene) and propane.

View **Project 1: HO2 Analysis**  [here](HTML/PR01-Rstartup-HO2.html). The code for this project is in the following RMD file. Data can be found in the data zip file [download from here](data/data.zip).

- [RMD](RMD/PR01-Rstartup-HO2.Rmd)

- [CSV](data/nyhh02.csv)

# Topic 2: Macro Financial Analysis {.tabset}

## Week 3

### View asynchronous material for Unit 3 -- Macrofinancial Analysis

We build on many of the finance concepts we covered in weeks 1 and 2. Here we begin to build out the 

- Stylized facts of the market including volatility clustering, leverage, and the persistence of returns and volatility 

- Visualize risk and return

Supporting files are in

- [RMD](RMD/W3-Macrofinancial.Rmd)

- [PDF](PDF/W3-Macrofinancial.pdf)

### Practice

For the adventurous among you try out the code used in the macro-financial asynch using data from the Fama and French market returns data set.

- [Fama and French stock market factors data set: fama-french-daily.csv](data/F-F_Research_Data_5_Factors_2x3_daily.csv)

- The factors in each column are defined [here](http://mba.tuck.dartmouth.edu/pages/faculty/ken.french/Data_Library/f-f_5developed.html)

- You can read more about this influential model from Fama and French [here](https://www8.gsb.columbia.edu/programs/sites/programs/files/finance/Finance%20Seminar/spring%202014/ken%20french.pdf)

### Workbook session

1. Welcome: meet and greet, respond to polls, list asynch highs and lows in chat room. (5 minutes)

2. The interactive workbook is [deployed here](https://wgfoote.shinyapps.io/finalytics-live-3/). Copy these files into your working directory to render your local version of this session:

- [Live Session #3 learnr RMD file](RMD/live-3.Rmd)

- [IYM data CSV file](RMD/IYM.csv)

- [USO data CSV file](RMD/USO.csv)

3. Review and discuss asynch questions and issues. Record solutions on the wall. (15 minutes)

4. Break-out #1: Review data inspection and exploration assignment together. Break out into separate rooms to discuss, analyze, and formulate report. Share report together. (30 minutes)

5. Break-out #2: Review autocorrelation assignment together. Break out into separate rooms to discuss, analyze, and formulate report. Share report together. (30 minutes)

6. Debrief this session and review next week's asynchronous materials. (10 minutes)

## Week 4

### View asynchronous material for Unit 3 - Macrofinancial Analysis again

We build on many of the finance concepts we covered in week 3. Here we begin to

- Relate markets to one another

- Visualize risk and return and correlation impacts

- Attempt to forecast GNP!

- Try to make our own inferential statistics using bootstrapping techniques

Supporting files continue to be in the `Unit 3` materials (even though we are in Week 4!)

- [RMD](RMD/W3-Macrofinancial.Rmd)

- [PDF](PDF/W3-Macrofinancial.pdf)

View a primer on [quantile regression here](HTML/primer-quantile-regression.html).

### Workbook session

1. Welcome: meet and greet, respond to polls, list asynch highs and lows in chat room. (5 minutes)

2. The interactive workbook is [deployed here](https://wgfoote.shinyapps.io/finalytics-live-4/). Copy this file into your working directory to render your local version of this session:

- [Workbook Session #4 learnr RMD file](RMD/live-4.Rmd)

3. Review and discuss asynch questions and issues. Record solutions on the wall. (15 minutes)

4. Assignment #1: Review cross correlation assignment together. Break out into separate rooms to discuss, analyze, and formulate report. Share report together. (30 minutes)

5. Assignment #2: Review market spillover assignment together. Break out into separate rooms to discuss, analyze, and formulate report. Share report together. (30 minutes)

6. Debrief this session and review next week's asynchronous materials. (10 minutes)

## Project 2: Foreign Exchange Markets {.tabset}

Live Sessions 3 and 4 will cover material to help successfully complete Project 2.

View a tutorial on **foreign exchange markets** [here](HTML/exchange-rates-1.html).

View **Project 2: foreign exchange market analysis**  [here](HTML/PR02_exrates-solution.html)

- [RMD](RMD/PR02_exrates-solution.Rmd)

# Topic 3: Market Risk {.tabset}

## Week 5

### View Market Risk asynchronous material deposited in Unit 4 -- Market Risk

Yes, we are in Week 5 by the calendar, and you will need to go to the Market Risk material that is located in Week 5. We happen to "synch" up!

- We will wander into various markets and repeat our macrofinancial data analysis. 

- Building on this, we will begin to look at the tails of the returns distributions and define risk measures

- You can practice this, and next week's, modeling efforts using the exchange rates, HO2, and other data sets you have already become very familiar with.

Supporting files (`Unit 4` materials)

- [RMD](RMD/W5-MarketRisk.Rmd)

- [PDF](PDF/W5-MarketRisk.pdf)

### Workbook session

1. Welcome: meet and greet, respond to polls, list asynch highs and lows in chat room. (5 minutes)

2. The interactive workbook is [deployed here](https://wgfoote.shinyapps.io/finalytics-live-5/). Copy this file into your working directory to render your local version of this session:

- [Workbook Session #5 learnr RMD file](RMD/live-5.Rmd)

3. Review and discuss asynch questions and issues. Record solutions on the wall. (15 minutes)

4. Assignment #1: Review GARCH simulation assignment together. Break out into separate rooms to discuss, analyze, and formulate report. Share report together. (30 minutes)

5. Assignment #2: Review historical simulation assignment together. Break out into separate rooms to discuss, analyze, and formulate report. Share report together. (30 minutes)

6. Debrief this session and review next week's asynchronous materials. (10 minutes)

## Week 6

This week we continue with

- Value at Risk (VaR) is a quantile and more importantly, the VaR of a combined portfolio can be larger than the sum of the VaRs of its components (sub-additivity). This makes VaR not very useful in developing estimates of capital needed to manage the risk of a business.

- Expected Shortfal (ES) is the loss in excess of a threshold. Thresholds are often the VaR quantile of loss in the tail of a loss distribution. The ESof a combined portfolio is indeed the sum of the ES of the components of portfolio. They are additive -- a very useful metric property for estimating capital, especially NPV changes.

- Again, for practice, use these metrics on prior weeks' data sets.

Supporting files continue to be (`Unit 4` materials)

- [RMD](RMD/W5-MarketRisk.Rmd)

- [PDF](PDF/W5-MarketRisk.pdf)

### Workbook session

1. Welcome: meet and greet, respond to polls, list asynch highs and lows in chat room. (5 minutes)

2. The interactive workbook is [deployed here](https://wgfoote.shinyapps.io/finalytics-live-6/). Copy this file into your working directory to render your local version of this session:

- [Workbook Session #6 learnr RMD file](RMD/live-6.Rmd)

3. Review and discuss asynch questions and issues. Record solutions on the wall. (15 minutes)

4. Assignment #1: Review  expected shortfall estimation assignment together. Break out into separate rooms to discuss, analyze, and formulate report. Share report together. (30 minutes)

5. Assignment #2: Review loss tail estimation assignment together. Break out into separate rooms to discuss, analyze, and formulate report. Share report together. (30 minutes)

6. Debrief this session and review next week's asynchronous materials. (10 minutes)

## Project 3: Metals Risk and Capital {.tabset}

Heavy metals

- Go to [this site to get some rock hard information](https://minerals.usgs.gov/minerals/pubs/commodity/) about nickel, copper, and aluminum (aka aluminium).

- The London Metals Exchange (LME) has a great [introductory presentation on metals hedging here](http://www.metalbulletin.com/events/download.ashx/document/speaker/7273/a0ID000000X0k1JMAR/Presentation) and [more general hedging information here](https://www.lme.com/en-GB/Trading/Physical-market-services/Hedging#tabIndex=1).

View these two versions of Project 3: Escapades in market risk analysis and a `flexdashboard` project example:

- [This version closely follows the R notation used in the asynchronous portion of the course.](HTML/PR03_market-risk.html) with [RMD file](RMD/PR03_market-risk.Rmd)

- [This version follows the development of the same analysis using tidyverse techniques from the synchronous workbook sessions](HTML/PR03_market-risk-rev.html) with [RMD file](RMD/PR03_market-risk-rev.Rmd)

- Here is an [example](HTML/ofcourse-42.html) of the use of `flexdashboard` for this project with [RMD file](RMD/ofcourse-42.Rmd).


# Topic 4: Portfolio Analytics {.tabset}

## Week 7

### View asynchronous material assigned to Unit 5 -- Portfolio Analytics

This week we travel into the 2 asset case

- We define risk tolerance (remember our extreme value work) and a threshold for loss

- We then specify a statistical confidence interval to build upper and lower bounds on our choices

- Then we use some very interesting math to solve for the optimal allocation of cash and risky asset.

Supporting files are in

Supporting files (`Unit 5` materials for use in this week 7)

- [RMD](RMD/W9-PortfolioAnalytics.Rmd)

- [PDF](PDF/W9-PortfolioAnalytics.pdf)

### Workbook session

1. Welcome: meet and greet, respond to polls, list asynch highs and lows in chat room. (5 minutes)

2. The interactive workbook is [deployed here](https://wgfoote.shinyapps.io/finalytics-live-7/). Copy this file into your working directory to render your local version of this session:

- [Workbook Session #7 learnr RMD file](RMD/live-7.Rmd)

3. Review and discuss asynch questions and issues. Record solutions on the wall. (15 minutes)

4. Assignment #1: Review portfolio construcion assignment together. Break out into separate rooms to discuss, analyze, and formulate report. Share report together. (30 minutes)

5. Assignment #2: Review risky asset-cash decision simulation assignment together. Break out into separate rooms to discuss, analyze, and formulate report. Share report together. (30 minutes)

6. Debrief this session and review next week's asynchronous materials. (10 minutes)

## Week 8

This week we continue into the multiple risky asset case

- Some matrix math helps us with describing and specifying portfolios

- We use a [quadratic programming solver](https://cran.r-project.org/web/packages/quadprog/quadprog.pdf) in R very similar to the capability of Frontline's Excel add-in solver.

- We then try to run the analysis full circle to use the optimal (tangency) risk asseet portfolio in the Week 7 model of cash and risky asset.

- The **risky asset** is really a portfolio that might be traded. In fact, this is what an [ETF (exchange-traded fund)](https://www.nasdaq.com/etfs/what-are-ETFs.aspx) is.

Supporting files (`Unit 5` materials) continue to be

- [RMD](RMD/W9-PortfolioAnalytics.Rmd)

- [PDF](PDF/W9-PortfolioAnalytics.pdf)

### Workbook session

1. Welcome: meet and greet, respond to polls, list asynch highs and lows in chat room. (5 minutes)

2. The interactive workbook is [deployed here](https://wgfoote.shinyapps.io/finalytics-live-8/). Copy this file into your working directory to render your local version of this session:

- [Workbook Session #8 learnr RMD file](RMD/live-8.Rmd)

3. Review and discuss asynch questions and issues. Record solutions on the wall. (15 minutes)

4. Assignment #1: Review risky portfolio allocation together. Break out into separate rooms to discuss, analyze, and formulate report. Share report together. (30 minutes)

5. Assignment #2: Review risky asset-cash decision (again) simulation assignment together. Break out into separate rooms to discuss, analyze, and formulate report. Share report together. (30 minutes)

6. Debrief this session and review next week's asynchronous materials. (10 minutes)

## Project 4: Metals Risk-Return Allocation

We continue the Ni-Cu-Al metals discussion with the computation and interpretation of the optimal allocation of these metals in a revenue portfolio.

- Be sure to note how we insert (slyly!) extreme value finance and confidence intervals through the use of a subsetting of the data by quantile (VaR) threshold and by sampling returns.

- The last question is of great import: what is the optimal allocation of optimal risk assets (Ni-Cu-Al) and cash (and high quality marketable securities)? The answer relies on the first part of the asynchronous material about the Iberian contract.

View [Project 4: metals portfolio analysis here](HTML/PR04_portfolio-analytics-rev.html)

- [RMD](RMD/PR04_portfolio-analytics-rev.Rmd)

# Topic 5: Enterprise Risk Analytics {.tabset}

## Week 9

### View asynchronous material for Unit 6 -- Enterprise Risk Analytics

- This is a capstone chunk of material that forces us to simulate loss for a portfolio

- We use a monte carlo technique called a copula to jointly generate portfolio returns

- We also allow ourselves the luxury of using non-normal distributions, like our old friend the `gamma` distribution.

- Applying this material to the metals data might prove very useful in understanding the level and the confidence intervals surrounding loss and capital.

Supporting files (`Unit 6` materials) are

- [RMD](RMD/W10-Aggregation.Rmd)

- [PDF](PDF/W10-Aggregation.pdf)

- [R](RMD/W10-ERM_appSHINY.R) file for the Shiny app in the asynchronous material

### Workbook session

1. Welcome: meet and greet, respond to polls, list asynch highs and lows in chat room. (5 minutes)

2. The interactive workbook is [deployed here](https://wgfoote.shinyapps.io/finalytics-live-9/). Copy this file into your working directory to render your local version of this session:

- [Workbook Session #9 learnr RMD file](RMD/live-9.Rmd)

3. Review and discuss asynch questions and issues. Record solutions on the wall. (15 minutes)

4. Assignment #1: Review monte carlo simulation of correlated innovations together. Break out into separate rooms to discuss, analyze, and formulate report. Share report together. (30 minutes)

5. Assignment #2: Review margin simulation assignment together. Break out into separate rooms to discuss, analyze, and formulate report. Share report together. (30 minutes)

6. Debrief this session and review next week's asynchronous materials. (10 minutes)

## Week 10

This last week finishes your preparation for your individual final project.

- You can use this [example of a financial web application](https://wgfoote.shinyapps.io/extreme-4/) built with the tools learned in this course as a template

- To answer all of the questions in the final project, you might need to increment the features of the example application.

### Workbook session

1. Welcome: meet and greet, respond to polls, list course highs and lows in chat room. (5 minutes)

2. The workbook is [deployed here](RMD/live-10.html). Copy this file into your working directory to render your local version of this session:

- [Workbook Session #10 flexdashboard RMD file](RMD/live-10.Rmd)

3. Review and discuss final project questions and issues. Record solutions on the wall. (15 minutes)

4. Assignment #1: Review choices of business contexts, data, analytical approach together. Break out into separate rooms to discuss, analyze, and formulate report. Share report together. (30 minutes)

5. Assignment #2: Review sample flexdashboard final project template together. Break out into separate rooms to discuss, analyze, and formulate report. Share report together. (30 minutes)

6. Debrief this session. (10 minutes)

## Project 5 -- FINAL

### This is it -- the FINALE

We have built everything from net present value and internal rate of return calculators to extreme market event indicators, optimal portfolios: all set against market risk and return stylized facts. The final project puts it all together for a business context and data set of your choice. 

### Assignment

This assignment will be the focus of Live Session 10. The final project is due at the Final Project Session in week 11. Submit into **Coursework > Assignments and Grading > Project 5 > Submission** an `RMD`  file with filename **lastname-firstname_Final-Project.Rmd** and associated `csv` or other relevant files. 

1. Choose a business decision and context. Based on that decision and context choose at least three (3) **daily** time series of sufficient length to help (often not completely) answer the key business questions that arise out of the business context and decision.

2. Build a flexdashboard application that addresses the business context you have chosen. The following outline is a guideline for the work flow in the web application. Populate the application with appropriate models, descriptive, explanatory, and predictive text, tables, and plots.

3. List in the text the `R` packages and skills needed to complete this project along with references used, including data sources.

4. Discuss how well did the results begin to answer the business questions posed for the project.

View [Project 5: FINAL here](HTML/PR05-FINAL.html)

- [RMD](RMD/PR05-FINAL.Rmd)

